---
title: "Automatic Grouping of Factors"
author: "Zining Qi"
output: workflowr::wflow_html
---

This analysis applies some automatic grouping of factors for the autoimmune GWAS data, based on the grouping information.

The grouping methods are:

1. A Simple Automatic Approach Based on ANOVA

2. Method from GCBD

The grouping information are:

1. Traits

2. Affected system

### Setup

Load required packages and data:

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#",
  collapse = TRUE,
  results = "hold",
  fig.align = "center",
  dpi = 120
)
```

```{r load-packages-data, message=FALSE}
# Data manipulation and visualization
library(tidyverse)
library(reshape2)
library(cowplot)
library(viridis)
library(Matrix)
library(ggplot2)

# Matrix factorization tools
library(flashier)
library(fastTopics)

# Load preprocessed GWAS effect size matrix
source('../code/group_factors.R')
load('../output/immune_mf_factors.RData')
load('../output/immune_snmf_factors.RData')
```

### A Simple Automatic Approach Based on ANOVA

#### flashier semi-NMF

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
F <- fl_snmf_ldf$F
k <- ncol(F)
colnames(F) <- paste0("k",1:k)

traits <- rownames(F)
affected_system <- c(
  "gastrointestinal", "Skin", "joints_muscles", 
  "spine_sacroiliac_joints", "nervous", 
  "gastrointestinal", "gastrointestinal", 
  "gastrointestinal", "multiple", 
  "nervous", "endocrine_glands"
)
```

Take a look at the elbow plot:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
ordered_df_traits <- ANOVA_factors(F, traits, stats = "R2")
ordered_df_system <- ANOVA_factors(F, affected_system, stats = "R2")
par(mfrow = c(2,1))
plot(ordered_df_traits$rank, ordered_df_traits$stats, type = "o", xlab = "Rank", ylab = "R2", main = "Traits", ylim = c(0,1.1))
text(ordered_df_traits$rank, ordered_df_traits$stats, labels = ordered_df_traits$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
plot(ordered_df_system$rank, ordered_df_system$stats, type = "o", xlab = "Rank", ylab = "R2", main = "Afftected System", ylim = c(0,1.1))
text(ordered_df_system$rank, ordered_df_system$stats, labels = ordered_df_system$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
par(mfrow = c(1,1))
```


From the plot, each factor is perfectly associated with traits. For affected system, k5, k4, k1, and k8 are perfectly associated with affected system, whereas k2, k6, and k7 have moderate association with affected system. 

Now, plot the structure plot by cutting off the factors with $R^2$ less than 0.7

```{r affected-system-0.7}
source('../code/plot_factors.R')

anno_data <- plot_factor_annotation_data(F, affected_system)
factors_to_plot <- paste0("k", ordered_df_system$factor[ordered_df_system$stats > 0.7])
F_ordered <- anno_data$F_ordered[, factors_to_plot]
# Create heatmap
p_snmf_grouping1 <- pheatmap::pheatmap(
  F_ordered,
  annotation_row = anno_data$annotation_df,
  annotation_colors = anno_data$annotation_colors,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = 'none',
  show_colnames = TRUE,
  border_color = "white",
  main = "Semi-NMF Factors by Affected Biological System",
  color = colorRampPalette(c("white", "#21908CFF"))(100)
)

p_snmf_grouping1
```


Then, plot the structure plot by cutting off the factors with $R^2$ less than 0.5.

```{r affected-system-0.7}
factors_to_plot <- paste0("k", ordered_df_system$factor[ordered_df_system$stats > 0.5])
F_ordered <- anno_data$F_ordered[, factors_to_plot]
# Create heatmap
p_snmf_grouping2 <- pheatmap::pheatmap(
  F_ordered,
  annotation_row = anno_data$annotation_df,
  annotation_colors = anno_data$annotation_colors,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = 'none',
  show_colnames = TRUE,
  border_color = "white",
  main = "Semi-NMF Factors by Affected Biological System",
  color = colorRampPalette(c("white", "#21908CFF"))(100)
)

p_snmf_grouping2
```


Also, plot the structure plot by cutting off the factors with $R^2$ less than 0.3.

```{r affected-system-0.7}
factors_to_plot <- paste0("k", ordered_df_system$factor[ordered_df_system$stats > 0.3])
F_ordered <- anno_data$F_ordered[, factors_to_plot]
# Create heatmap
p_snmf_grouping3 <- pheatmap::pheatmap(
  F_ordered,
  annotation_row = anno_data$annotation_df,
  annotation_colors = anno_data$annotation_colors,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = 'none',
  show_colnames = TRUE,
  border_color = "white",
  main = "Semi-NMF Factors by Affected Biological System",
  color = colorRampPalette(c("white", "#21908CFF"))(100)
)

p_snmf_grouping3
```


#### flashier Laplace MF 

Similary, we will try grouping factors by using this ANOVA method. 


```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
F <- fl_mf_ldf$F
k <- ncol(F)
colnames(F) <- paste0("k",1:k)

traits <- rownames(F)
affected_system <- c(
  "gastrointestinal", "Skin", "joints_muscles", 
  "spine_sacroiliac_joints", "nervous", 
  "gastrointestinal", "gastrointestinal", 
  "gastrointestinal", "multiple", 
  "nervous", "endocrine_glands"
)
```

Take a look at the elbow plot:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
ordered_df_traits <- ANOVA_factors(F, traits, stats = "R2")
ordered_df_system <- ANOVA_factors(F, affected_system, stats = "R2")
par(mfrow = c(2,1))
plot(ordered_df_traits$rank, ordered_df_traits$stats, type = "o", xlab = "Rank", ylab = "R2", main = "Traits", ylim = c(0,1.1))
text(ordered_df_traits$rank, ordered_df_traits$stats, labels = ordered_df_traits$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
plot(ordered_df_system$rank, ordered_df_system$stats, type = "o", xlab = "Rank", ylab = "R2", main = "Afftected System", ylim = c(0,1.1))
text(ordered_df_system$rank, ordered_df_system$stats, labels = ordered_df_system$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
par(mfrow = c(1,1))
```


From the plot, each factor is perfectly associated with traits. For affected system, k5, k2, k1, and k8 are perfectly associated with affected system, whereas k3, k7, and k6 have moderate association with affected system. 

Now, plot the structure plot by cutting off the factors with $R^2$ less than 0.7

```{r affected-system-0.7}
anno_data <- plot_factor_annotation_data(F, affected_system)
factors_to_plot <- paste0("k", ordered_df_system$factor[ordered_df_system$stats > 0.7])
F_ordered <- anno_data$F_ordered[, factors_to_plot]
# Create heatmap
p_mf_grouping1 <- pheatmap::pheatmap(
  F_ordered,
  annotation_row = anno_data$annotation_df,
  annotation_colors = anno_data$annotation_colors,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = 'none',
  show_colnames = TRUE,
  border_color = "white",
  main = "MF Factors by Affected System",
  color = colorRampPalette(c("#440154FF", "white", "#21908CFF"))(100)
)

p_mf_grouping1
```


Then, plot the structure plot by cutting off the factors with $R^2$ less than 0.5.

```{r affected-system-0.7}
factors_to_plot <- paste0("k", ordered_df_system$factor[ordered_df_system$stats > 0.5])
F_ordered <- anno_data$F_ordered[, factors_to_plot]
# Create heatmap
p_mf_grouping2 <- pheatmap::pheatmap(
  F_ordered,
  annotation_row = anno_data$annotation_df,
  annotation_colors = anno_data$annotation_colors,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = 'none',
  show_colnames = TRUE,
  border_color = "white",
  main = "MF Factors by Affected System",
  color = colorRampPalette(c("#440154FF", "white", "#21908CFF"))(100)
)

p_mf_grouping2
```


Also, plot the structure plot by cutting off the factors with $R^2$ less than 0.3.

```{r affected-system-0.7}
factors_to_plot <- paste0("k", ordered_df_system$factor[ordered_df_system$stats > 0.3])
F_ordered <- anno_data$F_ordered[, factors_to_plot]
# Create heatmap
p_mf_grouping3 <- pheatmap::pheatmap(
  F_ordered,
  annotation_row = anno_data$annotation_df,
  annotation_colors = anno_data$annotation_colors,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  scale = 'none',
  show_colnames = TRUE,
  border_color = "white",
  main = "MF Factors by Affected System",
  color = colorRampPalette(c("#440154FF", "white", "#21908CFF"))(100)
)

p_mf_grouping3
```






