---
title: "Exploring loadings of the autoimmune GWAS data"
author: "Zining Qi"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: console
---

This page is to check individual example for the preliminary results to see if results make sense. For example, we could pick out some variants have high loadings that sharing among traits. And then make forest plot for effect size and standard error for each trait.

```{r setup, include=FALSE}
# Set knitr options for consistent output formatting
knitr::opts_chunk$set(
  comment = "#",
  collapse = TRUE,
  results = "hold",
  fig.align = "center",
  dpi = 120
)
```

```{r load-packages, message=FALSE}
# Load required packages
library(tidyverse)    # For data manipulation and visualization
library(data.table)   # For efficient data reading and processing
library(pheatmap)
```

Load flash objects which have loadings and factors and dataframe with effect size and standard error of variants. 
```{r}
load('../data/autoimmune_combined_results.RData')
load('../output/immune_snmf_factors.RData')
```


Extract high loading variants for each loadings

```{r}
L <- fl_snmf_ldf$L
colnames(L) <- c(paste0('k', 1:ncol(L)))

L_df <- as.data.frame(L) %>%
  rownames_to_column("variant_id") %>%
  pivot_longer(cols = -variant_id, 
               names_to = "factor", 
               values_to = "loading")

merged_data <- combined_results %>%
  inner_join(L_df, by = "variant_id")

# Get top N variants per factor
top_variants_per_factor <- L_df %>%
  group_by(factor) %>%
  arrange(desc(abs(loading))) %>%
  slice_head(n = 10) %>%  # Top 5 variants per factor
  ungroup()
```


```{r}
# Get top 5 variants per factor
top_variants <- L_df %>%
  group_by(factor) %>%
  arrange(desc(abs(loading))) %>%
  slice(1:5) %>%  # Top 5 variants per factor
  ungroup()

# Merge with effect data
plot_data <- combined_results %>%
  inner_join(top_variants, by = "variant_id") %>%
  mutate(
    ci_low = beta - 1.96 * standard_error,
    ci_high = beta + 1.96 * standard_error,
    # Label for the variant (shows loading value)
    variant_label = paste0(variant_id, " (λ=", round(loading, 2), ")")
  )
```


```{r}
plot_single_variant <- function(variant_data) {
  # Prepare labels
  label_text <- cbind(
    c("Trait", variant_data$disease),
    c("Effect [95% CI]", 
      sprintf("%.2f [%.2f, %.2f]", 
              variant_data$beta, 
              variant_data$ci_low, 
              variant_data$ci_high))
  )
  
  forestplot(
    labeltext = label_text,
    mean = c(NA, variant_data$beta),
    lower = c(NA, variant_data$ci_low),
    upper = c(NA, variant_data$ci_high),
    title = paste("Variant:", variant_data$variant_label[1], 
                 "\nFactor:", variant_data$factor[1]),
    xticks = seq(round(min(variant_data$ci_low))-0.1, round(max(variant_data$ci_high))+0.1, 0.05),
    boxsize = 0.2,
    lineheight = unit(0.8, "cm"),
    col = fpColors(box = "royalblue", line = "darkblue", summary = "royalblue"),
    graphwidth = unit(6, "cm"),
    clip = c(-1, 1)  # Force consistent x-axis limits
  )
}

nested_plots <- plot_data %>%
  group_by(factor, variant_id) %>%
  group_split()

plot_single_variant(nested_plots[[1]])

all_plots <- map(nested_plots, plot_single_variant)

walk2(all_plots, nested_plots, ~{
  factor_dir <- file.path("../output/forest_plots", unique(.y$factor))
  if (!dir.exists(factor_dir)) dir.create(factor_dir, recursive = TRUE)
  
  png(file.path(factor_dir, paste0(unique(.y$variant_id), ".png")),
      width = 1200, height = 800, res = 150)
  print(.x)
  dev.off()
})
```








